---
title: Time Series Forecasting Practice
author: Will Bidstrup
date: '2018-07-03'
slug: time-series-forecasting-practice
categories:
  - R
tags:
  - forecasting
---

Today I explore time series forecasting use AFL data from the fitzRoy package.     

# Key concepts

A forecast is the mean of simulated futures.  

- Making a ts object  
- Trend, Seasonality, Cyclicity  
- Plotting lags and ACF  
- White noise  
- Creating forecast objects  
- Checking residuals
- Mean Absolute Scaled Error
- Cross Validation
- Exponential Smoothing

# Things to do

```{r}
# TODO - figure out why full ts doesn't show info from the 1900s
```


```{r, include = FALSE}
install.packages("tidyverse", repos="http://cran.rstudio.com/")
install.packages("here", repos="http://cran.rstudio.com/")
install.packages("ggthemes", repos="http://cran.rstudio.com/")
install.packages("forecast", repos="http://cran.rstudio.com/")
install.packages("padr", repos="http://cran.rstudio.com/")


install.packages("devtools", repos="http://cran.rstudio.com/")
devtools::install_github("jimmyday12/fitzRoy")


```

```{r, message = FALSE}
# Packages
library(tidyverse) # For everything
library(here) # For project-oriented workflow
library(ggthemes) # For nice plot themes
library(forecast) # For nice plot themes
library(devtools) # For non-CRAN packages
library(fitzRoy) # Data source
library(padr) # Dates padding

```

# Create a time series

```{r}
# First load results info  
d <- get_match_results()
```

Create a df with variables per time (mean where more than one game played on one day).  
```{r}
# Then prepare a subset for ts analysis 
d_prep_ts <- d %>%
  group_by(Date) %>%
  summarize(Margin = abs(mean(Margin, na.rm = TRUE)),
            Total_Points = mean((Home.Points + Away.Points), na.rm = TRUE),
            Conversion = mean((Home.Goals + Away.Goals) / (Home.Goals + Away.Goals + Home.Behinds + Away.Behinds)))


```

The [padr package](https://edwinth.github.io/blog/padr-intro/) looks very helpful to fill in missing days.

```{r}
# Fill timeseries with missing dates
d_prep_ts_full <- d_prep_ts %>% pad
  
```


```{r}
# TODO - how to add missing weeks for full year coverage?
# Create ts object
d_ts <- ts(d_prep_ts_full[, 2:4], start = c(1897-05-08), frequency = 365.25) # Arbitrarily screwing with the frequency here
```



Having issues visualizing the data pre 1960?  I suspect it's something to do with games only being played once a week.
```{r}
# TODO - why is data before 1960 not showing?
autoplot(d_ts, facets = TRUE)
```

The points are visible in a standard ggplot...
```{r}
ggplot(data = d_prep_ts_full, aes(x = Date, y = Margin)) +
  geom_point()

ggplot(data = d_prep_ts_full, aes(x = Date, y = Total_Points)) +
  geom_point()

ggplot(data = d_prep_ts_full, aes(x = Date, y = Conversion)) +
  geom_point()
```


Subset 1900 - 1950.

```{r}
nineteen_hundreds <- window(d_ts, 1900, 1950)
```

This makes no sense...
```{r}
# TODO - why is data before 1960 not showing?
autoplot(nineteen_hundreds, facets = TRUE)
```

Subset since 1980, this seems to work.  
```{r}
my_life_afl <- window(d_ts, 1980)
```

```{r}
autoplot(my_life_afl, facets = TRUE)
```


1980 only
```{r}
birth_year <- window(d_ts, 1980, 1981)

```

# Lag plots

Create yearly timeseries.  

```{r}
d_yearly <- d %>%
  filter(Date > "1980-01-01" & Date < "1989-12-31")
```

```{r}
d_yearly <- d_yearly %>%
  separate(Date, c("Year", "Month", "Day"), sep = "-")
```

```{r}

d_yearly <- d_yearly %>%
  group_by(Year) %>%
  summarize(Margin = abs(mean(Margin, na.rm = TRUE)),
            Total_Points = mean((Home.Points + Away.Points), na.rm = TRUE),
            Conversion = mean((Home.Goals + Away.Goals) / (Home.Goals + Away.Goals + Home.Behinds + Away.Behinds)))

```

```{r}
d_yearly_ts <- ts(d_yearly[, 2:4], start = c(1980, 1), frequency = 1)
```

```{r, fig.width = 15, fig.height = 10}
# TODO - save image and call directly
gglagplot(d_yearly_ts)
```

```{r}
# Select only conversion
d_yearly_conversion <- d_yearly %>%
  select(Year, Conversion)

d_yearly_conversion_ts <- ts(d_yearly_conversion[, 2], start = c(1980, 1), frequency = 1)

```

```{r}
ggAcf(d_yearly_conversion_ts)
```


```{r}
# Select only Margin
d_yearly_margin <- d_yearly %>%
  select(Year, Margin)

d_yearly_margin_ts <- ts(d_yearly_margin[, 2], start = c(1980, 1), frequency = 1)

```

```{r}
ggAcf(d_yearly_margin_ts)
```

# White Noise

Use Ljung-Box test to see if there are any significant correlations, then look more closely at individual variable ACF



```{r}
# A random time-series
```


```{r}
Box.test(d_yearly_conversion_ts, lag = 9, type = "Ljung")
# Large p value confirms visual inspection
```


# Naive Forecast

Previous value predicts future.  A **seasonal naive** uses info from previous season.  

```{r}
fcd_yearly_margin_ts <- naive(d_yearly_margin_ts, 5)
```

```{r}
autoplot(fcd_yearly_margin_ts)
summary(fcd_yearly_margin_ts)
```


```{r}
fcd_yearly_conversion_ts <- naive(d_yearly_conversion_ts, 5)
```

```{r}
autoplot(fcd_yearly_conversion_ts)
summary(fcd_yearly_conversion_ts)
```

# Residuals

```{r}
checkresiduals(fcd_yearly_conversion_ts)
```


# Training and test

I'm going to try to forecast average total points per game based on data from 1897 until now.  This is a purely theoretical interest.  

```{r}
d_points <- d %>%
  filter(Date > "1897-01-01" & Date < "2018-07-01")

d_points <- d_points %>%
  separate(Date, c("Year", "Month", "Day"), sep = "-")
```

```{r}
d_points <- d_points %>%
  group_by(Year) %>%
  summarize(Mean_Total_Points = mean((Home.Points + Away.Points), na.rm = TRUE))
```


```{r}
# Make ts object
d_points_ts <- ts(d_points[, 2], start = c(1897,1), frequency = 1)

str(d_points_ts)
```


First a look at the data. 

```{r}
autoplot(d_points_ts)
summary(d_points_ts)
```


White noise test.

```{r}
Box.test(d_points_ts, lag = 100, type = "Ljung") # low p value
```



Now split into test and train  

```{r}
train <- window(d_points_ts, end = 1995) #Use approx 80% of years as training
test <- window(d_points_ts, start = 1996) 
```



Create forecast on the train  

```{r}
# Naive forecast
naive_fc <- naive(train, h = 22)
```

```{r}
# Forecasts the mean! A good comparison point
mean_fc <- meanf(train, h = 22)
```


Check residuals  

```{r}
checkresiduals(naive_fc) # Looks like there is a pattern in residuals, and not normally distrubuted
```

```{r}
checkresiduals(mean_fc) # A different pattern
```


Plot with actual test results overlayed  

```{r}

autoplot(naive_fc) +
  autolayer(test, series = "Test data")
```


```{r}

autoplot(mean_fc) +
  autolayer(test, series = "Test data")
```


Compare accuracy

```{r}
accuracy(naive_fc, test)
```

```{r}
accuracy(mean_fc, test) # Error values all much bigger than the naive method
```


Apply to future  

```{r}
# Apply naive forecast to future

future <- naive(d_points_ts, h = 10)

```


```{r}
autoplot(future)
summary(future) # Highs getter higher, lows get lower, further into future
```



# Cross validation

Cross validation 8 times using `tsCV`. Useful for smaller smaples where train/test not as useful.  

```{r}
# TODO - troubleshoot this
# Make an empty matrix
#e <- matrix(NA_real_, nrow = 122, ncol = 8)
```

```{r}
# Fill empty matrix with forecast based on tsCV function
#for (h in 1:8)
  #e[, h] <- tsCV(d_points_ts, naive, h = h)
```


```{r}
#d_points_ts

```

# Simple Exponential Smoothing


9 possible exponential smoothing methods and 2 error methods so 18 models

SES method on the points time series.  

```{r}
# SES forecast
ses_fc <- ses(train, h = 22)
```

```{r}
summary(ses_fc)
```

```{r}
accuracy(ses_fc, test) # Not as good as naive in this case!

checkresiduals(ses_fc)

# TODO - full accuracy assessment
```


```{r}
autoplot(ses_fc) +
  autolayer(fitted(ses_fc))
```

Adding trends.  

SES OK if there is no trend or seasonality.  

Holt adds linear trends. Damped trend levels off.  

```{r}
fc_holt <- holt(train, h = 22)
```

```{r}
summary(fc_holt)
```

```{r}
accuracy(fc_holt, test)
```


```{r}
checkresiduals(fc_holt)
```

```{r}
autoplot(fc_holt) +
  autolayer(test, series = "Test Series")
```

Holt-Winters can account for trend and seasonality 

# State Space Models

Create a model then pass that to forecast which automatically picks best forecast method to minimize error.

The model state needs to be defined in terms of Error, Trend and Seasonality (ETS).

Great for getting quick results.  


```{r}
# Create ETS model (default)
fit_train <- ets(train)
```

```{r}
checkresiduals(fit_train)
```


```{r}
autoplot(forecast(fit_train))
summary(forecast(fit_train))
accuracy(forecast(fit_train), test)
```

# Transformations

Lambda.

```{r}
BoxCox.lambda(d_points_ts)
```


For non-seasonal differentiation, case use `diff()` - can also take seasonal difference as well as lag-1.  

```{r}
autoplot(d_points_ts)
```

```{r}
autoplot(diff(d_points_ts))
```

```{r}
ggAcf(diff(d_points_ts))
```


# ARIMA

AutoRegressive Integrated Moving Average models. 

AR - multiple regression with lagged observations as predictors.  

MA - multiple regression with lagged errors as predictors.  

Can only work with stationary data (so diffrerence if not already stationary)  

ARIMA(p, d, q)  

p = number of ordinary AR (observation) lags
d = number of lag-1 differences
q = number of ordinary MA (error) lags

*Can't compare ARIMA AICc between ARIMA and ETS* Us tsCV to do this.  


```{r}
# Create ARIMA model (NOTE can also set lambda)
fit_arima_train <- auto.arima(train)

```

```{r}
checkresiduals(fit_arima_train)
```

```{r}
summary(fit_arima_train)
```


```{r}
# Create forecast from ARIMA model

fc_arima <- forecast(fit_arima_train, h = 22)
```


```{r}
# Check accuracy
accuracy(fc_arima, test)

```

```{r}
# Show performance against test
autoplot(fc_arima) +
  autolayer(test, series = "Test Series")
```

```{r}
# Apply forecast to future 10 years with AUTO ARIMA
fit_arima_future <- auto.arima(d_points_ts)

fc_arima_future <- forecast(fit_arima_future, h = 10)

autoplot(fc_arima_future) +
  autolayer(test, series = "Test Series")

summary(fc_arima_future)

```



```{r}
# Try tweaking the ARIMA (include drift)
d_points_ts %>% 
  Arima(order = c(0,1,1), include.constant = TRUE) %>% 
  forecast(h = 10) %>%
  autoplot()
  
```

```{r}
# TODO - create a seasonal ts and try an ARIMA model on it
```


# Dynamic Regression

Can add matrix of predictor variable to auto.arima model.  

Might be interesting to look at forecasting total points informed by kicks.   

Dynamic harmonic regression incorporates seasonality with Fourier terms.  

# TBATS

Trigonometric terms for seasonality

Box-Cox

ARMA error

Trend

Seasonal  

Useful but dangerous!!  