---
title: EDA with R and PostgreSQL
author: Will Bidstrup
date: '2018-08-27'
slug: eda-with-r-and-postgresql
categories:
  - R
  - SQL
tags:
  - R
  - SQL
  - EDA
---

Today I connect my R session to a PostgreSQL database.  

I'm using the tutorial database of DVD rentals and [this](https://www.r-bloggers.com/getting-started-with-postgresql-in-r/). 

My objectives are to;  

- learn how to connect R to a database  
- practice using SQL queries within R Markdown  
- uncover material to use in the SQL project for the Udacity Programming For Data Science Nanodegree


```{r, include = FALSE}
install.packages("tidyverse", repos="http://cran.rstudio.com/")
install.packages("here", repos="http://cran.rstudio.com/")
install.packages("ggthemes", repos="http://cran.rstudio.com/")
install.packages("RPostgreSQL", repos="http://cran.rstudio.com/")
```

```{r, message = FALSE}
library(tidyverse)
library(here)
library(ggthemes)
library(RPostgreSQL)
```

```{r, include = FALSE}
pw <- "elitecassowary77"
```

```{r}
drv <- dbDriver("PostgreSQL")
```


```{r}
con <- dbConnect(drv, dbname = "dvdrental",
                 host = "localhost", port = 5432,
                 user = "postgres", password = pw)
```


```{r, include = FALSE}
rm(pw)
```


# Exploring the database

There are 15 tables in this relational database, connected as follows;  

*Insert pic of ERD*


# Question 1

```{r}
# Get results from db via sql query
q1 <- dbGetQuery(con, "WITH t1 AS
(SELECT *
FROM film f
LEFT JOIN film_category fc
ON f.film_id = fc.film_id
LEFT JOIN category c
ON c.category_id = fc.category_id
LEFT JOIN inventory i
ON i.film_id = f.film_id
LEFT JOIN rental r
ON r.inventory_id = i.inventory_id)

SELECT DISTINCT(t1.title) film_title,
        t1.name category_name,
        CASE WHEN t1.name IN ('Animation', 'Children', 'Classics', 'Comedy', 'Family', 'Music')
        THEN 'FAMILY' ELSE 'ADULT ONLY' END AS family_or_not,
        COUNT(t1.rental_id) OVER (PARTITION BY t1.title) AS rental_count
FROM t1
ORDER BY t1.name, t1.title;")

```


```{r}
# Create a plot

ggplot(data = q1 %>%
         group_by(category_name, family_or_not) %>%
         summarise(rental_count = sum(rental_count, na.rm = T)), aes(x = reorder(category_name, rental_count), y = rental_count)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap(. ~ family_or_not) +
  theme_tufte() +
  labs(title = "Adults like Sports, Families like Animation",
       subtitle = "",
       x = "Film category",
       y = "Count of rentals")

ggsave("q1.pdf")
```

```{r}
# Distribution of rental by category
ggplot(data = q1, aes(x = rental_count, fill = category_name)) +
  geom_density(alpha = 0.2) +
  theme_tufte() +
  facet_wrap(. ~ family_or_not) +
  theme_tufte() 

```

```{r}
# Distribution of rental by category
ggplot(data = q1, aes(x = rental_count, fill = family_or_not)) +
  geom_density(alpha = 0.2) +
  theme_tufte() 
```



# Question 1b

```{r}
# Get results from db via sql query
q1b <- dbGetQuery(con, "WITH t1 AS
(SELECT *
FROM film f
LEFT JOIN film_category fc
ON f.film_id = fc.film_id
LEFT JOIN category c
ON c.category_id = fc.category_id
LEFT JOIN inventory i
ON i.film_id = f.film_id
LEFT JOIN rental r
ON r.inventory_id = i.inventory_id)

SELECT DISTINCT(t1.title) film_title,
        t1.name category_name,
        t1.length,
        CASE WHEN t1.name IN ('Animation', 'Children', 'Classics', 'Comedy', 'Family', 'Music')
        THEN 'FAMILY' ELSE 'ADULT ONLY' END AS family_or_not,
        COUNT(t1.rental_id) OVER (PARTITION BY t1.title) AS rental_count
FROM t1
ORDER BY t1.name, t1.title;")

```


```{r}
ggplot(data = q1b, aes(x = length, y = rental_count)) +
  geom_point(alpha = 0.2)

length_matters <- cor(q1b$length, q1b$rental_count)

```



```{r}
# 
lm_1 <- lm(q1b$rental_count ~ q1b$length)

test <- coefficients(lm_1)

ggplot(data = q1b, aes(x = length, y = rental_count)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  facet_wrap(. ~ family_or_not) +
  labs(title = "",
       subtitle = "",
       x = "Length of film",
       y = "Count of rentals") +
  theme_minimal() +
  theme(text = element_text(size=24))
  




ggsave("q1b.pdf")


```



# Question 2


```{r}
q2 <- dbGetQuery(con, "WITH t1 AS
(SELECT *
FROM film f
JOIN film_category fc
ON f.film_id = fc.film_id
JOIN category c
ON c.category_id = fc.category_id)

SELECT DISTINCT(t1.title) film_title,
        t1.name category_name,
        CASE WHEN t1.name IN ('Animation', 'Children', 'Classics', 'Comedy', 'Family', 'Music')
        THEN 'FAMILY' ELSE 'ADULT ONLY' END AS family_or_not,
        t1.rental_duration,
        NTILE(4) OVER (ORDER BY t1.rental_duration) AS standard_quartile
FROM t1
ORDER BY t1.name, t1.title;")


```

```{r}

ggplot(data = q2, aes(x = category_name, y = rental_duration)) +
  geom_boxplot() +
  coord_flip()

```


```{r}
ggplot(data = q2, aes(x = rental_duration, fill = family_or_not)) +
  geom_density(alpha = 0.2) +
  theme_tufte() 
```


# Question 2b


```{r}
q2b <- dbGetQuery(con, "WITH t1 AS
(SELECT *
FROM film f
JOIN film_category fc
ON f.film_id = fc.film_id
JOIN category c
ON c.category_id = fc.category_id)

SELECT DISTINCT(t1.title) film_title,
        t1.name category_name,
        t1.length,
        CASE WHEN t1.name IN ('Animation', 'Children', 'Classics', 'Comedy', 'Family', 'Music')
        THEN 'FAMILY' ELSE 'ADULT ONLY' END AS family_or_not,
        t1.rental_duration,
        NTILE(4) OVER (ORDER BY t1.rental_duration) AS standard_quartile
FROM t1
ORDER BY t1.name, t1.title;")


```

```{r}
ggplot(data = q2b, aes(x = length, y = standard_quartile)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  facet_grid(. ~family_or_not) +
  labs(title = "",
       subtitle = "",
       x = "Length of film",
       y = "Duration of rental (quartile)") +
  theme_minimal() +
  theme(text = element_text(size=24)) 

ggsave("q2b.pdf")
```





# Question 3


```{r}
q3 <- dbGetQuery(con, "WITH  t1 AS (
      SELECT *
      FROM film f
      LEFT JOIN film_category fc
      ON f.film_id = fc.film_id
      LEFT JOIN category c
      ON c.category_id = fc.category_id),


      t2 AS (
        SELECT DISTINCT(t1.name) category_name,
        CASE WHEN t1.name IN ('Animation', 'Children', 'Classics', 'Comedy', 'Family', 'Music')
        THEN 'FAMILY' ELSE 'ADULT ONLY' END AS family_or_not,
        NTILE(4) OVER (ORDER BY t1.rental_duration) AS standard_quartile,
        t1.title title
        FROM t1)


SELECT DISTINCT(t2.category_name),
        t2.family_or_not,
        t2.standard_quartile,
        COUNT(t2.title) OVER (PARTITION BY t2.category_name, t2.standard_quartile ORDER BY t2.standard_quartile)
FROM t2
ORDER BY 1, 3;")


```

```{r}

ggplot(data = q3, aes(x = standard_quartile, y = count, fill = category_name)) +
  geom_bar(stat = "identity") +
  theme_tufte() +
  facet_grid(. ~ family_or_not)

```

```{r}

ggplot(data = q3, aes(x = standard_quartile, y = count, fill = category_name)) +
  geom_density(stat = "identity", alpha = 0.8) +
  theme_tufte() +
  facet_grid(. ~ family_or_not) +
  labs(title ="",
       subtitle = "",
       x = "Duration of rental (quartile)",
       y = "Count of distinct film title") +
  theme_minimal() +
  theme(text = element_text(size=24))

ggsave("q3.pdf")

```




```{r}
q3_with_r <- dbGetQuery(con, "SELECT *
      FROM film f
      LEFT JOIN film_category fc
      ON f.film_id = fc.film_id
      LEFT JOIN category c
      ON c.category_id = fc.category_id")

names(q3_with_r)
  
  
```








# Question 4
```{r}
q4 <- dbGetQuery(con, "
WITH t1 AS (
SELECT DATE_PART('month',r.rental_date) AS rental_month,
      DATE_PART('year',r.rental_date) AS rental_year,
      sto.store_id store_id,
      r.rental_id
FROM store sto
JOIN staff sta
ON sta.store_id = sto.store_id
JOIN rental r
ON sta.staff_id = r.staff_id)

SELECT t1.rental_month,
      t1.rental_year,
      t1.store_id,
      COUNT(t1.rental_id) rental_count
FROM t1
GROUP BY 1, 2, 3
ORDER BY 1, 2;
")

sum(q4$rental_count)

```



```{r}
ggplot(data = q4, aes(x = rental_month, y = rental_count, fill = factor(store_id))) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(. ~ rental_year) +
  theme_minimal() +
  labs(title = "",
       subtitle = "",
       x = "Rental month",
       y = "Count of rentals")  +
  theme(text = element_text(size=24))

ggsave("q4.pdf")
```









# Check base tables

```{r}
rental <- dbGetQuery(con, "SELECT * FROM rental")
```


