---
title: python training first steps
author: Will Bidstrup
date: '2018-06-22'
slug: python-training-first-steps
categories:
  - Python
tags:
  - Python
  - visualisation
---

Today I train on Python skills.

All code is in R except where commented prior as "# Python code"

```{r, include = FALSE}
install.packages("reticulate", repos="http://cran.rstudio.com/")


```

```{r}
library(reticulate) # to enable python code chunks in R Markdown

```


# matplotlib

Matplotlib is a visualization package/library.  

```{python}
# Python code
# Import matplotlib
import matplotlib.pyplot as plt
```

Using example data
```{python}
# Python code
# Create some data
year = [1950, 1970, 1990, 2010]
pop = [2.519, 3.692, 5.263, 6.972]

```

```{python}
# Python code
# Line plot
plt.plot(year, pop)
plt.xlabel('Year')
plt.ylabel('Population')
plt.title('Population Projections')
plt.show()
plt.clf()
```

```{python}
# Python code
# Scatterplot
plt.scatter(year, pop)
plt.show()
plt.clf()
```

# conclusion

I tried to import a data set as csv and use pandas as well. What I found was that it's not possible to use R Markdown interactively with Python commands. I do want to get Python work published in this blog, to do so I will follow this approach;  

1) Learn Python in terminal and anaconda  
2) Figure out how to publish HTML jupyter notebooks  
3) Figure out how to deploy those notebooks into this website  

