---
title: 2018 Makeover Monday visualisations
author: 'Will Bidstrup'
date: '2018-04-16'
slug: makeover-monday-visualisations
categories:
  - R
tags:
  - R
  - visualisation
---
I use [Makeover Monday](http://www.makeovermonday.co.uk/) to try new visual techniques in R and Tableau.  

```{r, include = FALSE}
# Install and load packages
install.packages("tidyverse", repos="http://cran.rstudio.com/")
library(tidyverse)

install.packages("here", repos="http://cran.rstudio.com/")
library(here)

install.packages("ggthemes", repos="http://cran.rstudio.com/")
library(ggthemes)

install.packages("ggrepel", repos="http://cran.rstudio.com/")
library(ggrepel)

```

# Week 16 Zambia Southern Province Confirmed Malaria Cases

```{r, include = FALSE}
# Read data
d <- read_csv("data/W16_MM_malaria.csv")
```

```{r, include = FALSE}
glimpse(d)
```

```{r, include = FALSE}
d$Date <- as.Date(d$Date, format = "%d/%m/%Y")


```

```{r, include = FALSE}
plot(ecdf(d$Cases))
```


```{r, echo = FALSE}
ggplot(data = d, aes(x = Date, y = Cases, fill = District)) +
  geom_bar(stat = "identity", position = "fill") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme_tufte() +
  labs(title = "Zambia Southern Province Confirmed Malaria Cases", x = "Date" , y = "Proportion of Cases",
       subtitle = "Simulated data from http://visualizenomalaria.org.",
       caption = expression(paste(italic("Contact jdrummey@path.org for data questions"))))
  
```





# Week 15 Arctic Sea Ice

```{r, include = FALSE}
# Read data
d <- read_csv("data/W15_MM_arctic_sea_ice.csv")
```

```{r, include = FALSE}
# Convert string to date
d$Date <- as.Date(d$Date, format = "%d/%m/%Y")

# Create fulldate
d$full_date <- d$Date

# Seperate Date into parts
d2 <- separate(d, "Date", c("year", "Month", "day", sep = "-"))

d2$`-` <- NULL
```

```{r, echo = FALSE}
# Here is the code that makes the plot
ggplot(data = d2, aes(x = year, y = `Extent (million sq km)`)) +
    geom_boxplot(data = d2, aes(x = year, y = `Extent (million sq km)`)) +
  geom_point(data = d2, aes( col = Month),position = "jitter", alpha = 0.2) +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  +
  labs(title = "Arctic Sea Ice Coverage", x = "Year" , y = "Extent of coverage measured (million sq km)",
       subtitle = "Coverage distribution reducing over time",
       caption = expression(paste(italic("Source: National Snow & Ice Data Center"))))

```

[Tableau version](https://public.tableau.com/views/ArcticSeaIceMeasurementOverTime/Dashboard1?:embed=y&:display_count=yes&publish=yes).

# Week 14 Wine Production

New techniques used: `dplyr::lag` to add lag variables and calculate change over time and `scale_color_manual` for palette  

Feedback: Need to add some plain English description of what the chart is showing (the x variable is potentially confusing)    

```{r, include = FALSE}
# Read data
d <- read_csv("data/W14_MM_wine_production.csv")
```

```{r, include = FALSE}
# Add the 2016 vs 2015 variance called out in the reports
d2 <- d %>%
  group_by(Country) %>%
  mutate(var.volume = dplyr::lag(`Wine production in mhl`, n = 1, default = NA),
         var.percent = round((`Wine production in mhl` - var.volume) / `Wine production in mhl`, digits = 2))

# Add classification to EU vs rest of world
d2$group <- ifelse(d$Country %in% c("Italy", "France", "Spain", "Germany", 
                                    "Portugal", "Romania", "Greece", "Austria",
                                    "Hungary", "Moldova", "Bulgaria", "Georgia", "Switzerland"), "EU",
                   ifelse(d$Country == "World total", "World", "Non-EU"))
d2$group <- as.factor(d2$group)

# Make country a factor
d2$Country <- as.factor(d2$Country)  

# Rename variables
names(d2) <- c("Country", "Year", "Production", "var.volume", "var.percent", "group")

# Calculate average variance
d3 <- d2 %>%
  group_by(Country) %>%
  summarise(avg_var = round(mean(var.percent, na.rm = TRUE), digits = 2))


# Add average variance to long data set
d2 <- d2 %>% 
  left_join(d3, by = "Country")

# Calculate var distance from average
d2$var.from.avg <- d2$var.percent - d2$avg_var

# Add positive or negative factor (for colouring)
d2$`2016 vs 2015` <- ifelse(d2$var.percent >= 0, "Growing", "Declining")
d2$`2016 vs 2015` <- factor(d2$`2016 vs 2015`, levels = c("Growing", "Declining"))




# Add category

d2$type <- ifelse(d2$var.percent > 0.05 & d2$var.from.avg > 0.05, "Growing - better than average",
                  ifelse(d2$var.percent < 0.05 & d2$var.from.avg < 0.05, "Declining - worse than average",
                         ifelse(d2$var.percent > 0.05 & d2$var.from.avg < 0.05, "Growing - as expected", "Declining - as expected")))

```

```{r, echo = FALSE}
# Here is the code that generates the plot
ggplot(data = d2 %>%
         filter(Year == "2016", Country != "World total"), aes(x = var.from.avg, y = var.percent, size = Production, col = `2016 vs 2015`)) +
  geom_point(alpha = 0.5) +
 xlim(c(-1, 1)) + 
  ylim(c(-1.2, 1.2)) + 
  ggrepel::geom_label_repel(aes(label = Country), data = d2 %>%
         filter(Year == "2016", Country != "World total")) +
  facet_grid(~ group) +
  guides(size=FALSE) +
  theme_tufte() +
  scale_color_manual(values=c("#722f37","#91a3b0")) +
  labs(title = "Wine Production in 2016 - who is improving and by how much?",
       subtitle = "Good year for New Zealand, bad year for South America and Hungary",
       x = "Variance 2016 vs 2015 compared to average Y/Y variance since 2012",
       y = "Variance in Wine Production 2016 vs 2015",
       caption = expression(paste(italic("Source: International Organisation of Vine and Wine 2017"))))
```


# Week 13 UK Top Ten Chocolate Bars

```{r, include = FALSE}
# Read data
d <- read_csv("data/W13_MM_top_ten_chocolate_bars.csv")
```

```{r, include = FALSE}
d$Brand <- as.factor(d$Brand)
d$`Age Group` <- as.factor(d$`Age Group`)
```

```{r, include = FALSE}
d <- na.omit(d)
```


```{r, echo = FALSE}

ggplot(data = d, aes(x = reorder(Brand, -Rank), y = reorder(Rank, -Rank), col = factor(`Age Group`, levels = levels(`Age Group`)))) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  theme(axis.text.y = element_text(angle = 90, hjust = 1)) +
  facet_grid(~ `Age Group`) +
  theme_tufte() +
  theme(legend.position="none") +
  labs(title = "UK highest rated chocolate bars by age group", x = "Brand" , y = "Ranking (10 is low, 1 is high)",
       subtitle = "Mars and Galaxy popular for all ages",
       caption = expression(paste(italic("Source: CDA"))))
```




